\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2024-03-07 misc/JSS article/jss-article-rnw/article.tex
Adding \proglang for C/C++ etc.

\code{collapse::pivot} provides a modern class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings: it has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots, and supports complex labelled data without loss of information. Its syntax is:
%
\begin{Code}
pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
  how = "longer", na.rm = FALSE, factor = c("names", "labels"),
  check.dups = FALSE, nthreads = 1, fill = NULL, drop = TRUE,
  sort = FALSE, transpose = FALSE)
\end{Code}
%
The demonstration below employs a generated dataset about fruits. We could equivalently think about a survey with households and individuals, or sectors and firms. Variable labels are stored in \code{attr(column, "label")}. The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation} provides more elaborate examples.
%
\begin{Schunk}
\begin{Sinput}
R> data <- data.frame(type = rep(c("A", "B"), each = 2),
+    type_name = rep(c("Apples", "Bananas"), each = 2),
+    id = rep(1:2, 2), r = abs(rnorm(4)), h = abs(rnorm(4)*2))
R> setrelabel(data, id = "Fruit Id", r = "Fruit Radius", h = "Fruit Height")
R> print(data)
\end{Sinput}
\begin{Soutput}
  type type_name id      r     h
1    A    Apples  1 0.3916 1.845
2    A    Apples  2 1.0761 1.667
3    B   Bananas  1 0.4958 1.976
4    B   Bananas  2 1.2679 2.974
\end{Soutput}
\begin{Sinput}
R> vlabels(data)
\end{Sinput}
\begin{Soutput}
          type      type_name             id              r              h 
            NA             NA     "Fruit Id" "Fruit Radius" "Fruit Height" 
\end{Soutput}
\end{Schunk}
%
To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = c(...))}. If \code{labels = "lab_name"} is specified, variable labels are saved to an additional column named \code{lab\_name}. In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to assign alternative names to the \code{variable} and \code{value} columns, respectively.
%
\begin{Schunk}
\begin{Sinput}
R> (dl <- pivot(data, ids = c("type", "type_name", "id"), labels = "label"))
\end{Sinput}
\begin{Soutput}
  type type_name id variable        label  value
1    A    Apples  1        r Fruit Radius 0.3916
2    A    Apples  2        r Fruit Radius 1.0761
3    B   Bananas  1        r Fruit Radius 0.4958
4    B   Bananas  2        r Fruit Radius 1.2679
5    A    Apples  1        h Fruit Height 1.8454
6    A    Apples  2        h Fruit Height 1.6666
7    B   Bananas  1        h Fruit Height 1.9762
8    B   Bananas  2        h Fruit Height 2.9737
\end{Soutput}
\begin{Sinput}
R> vlabels(dl)
\end{Sinput}
\begin{Soutput}
      type  type_name         id   variable      label      value 
        NA         NA "Fruit Id"         NA         NA         NA 
\end{Soutput}
\end{Schunk}
%
\fct{pivot} only requires essential information and intelligently guesses the rest. For example, the same result could have been obtained by \code{pivot(data, values = c("r", "h"), labels = "label")}. An exact reverse operation can also be performed by specifying as little as \code{pivot(dl, labels = "label", how = "w")}. \newline

The second option is a wider pivot with \code{how = "wider"}. Here, \code{names} and \code{labels} can be used to select columns containing the names of new columns and their labels.\footnote{multiple columns with names and labels could be selected, which would be combined using \code{"\_"} for names and \code{" - "} for labels.} Note how the labels are combined with existing labels such that also this operation is without loss of information. It is, however, a destructive operation, i.e., with 2 or more columns selected through \code{values}, \fct{pivot} is not able to reverse it. Further arguments like \code{na.rm}, \code{fill}, \code{sort}, and \code{transpose} can be used to control the casting process.
%
\begin{Schunk}
\begin{Sinput}
R> (dw <- pivot(data, "id", names = "type", labels = "type_name", how = "w"))
\end{Sinput}
\begin{Soutput}
  id    r_A    r_B   h_A   h_B
1  1 0.3916 0.4958 1.845 1.976
2  2 1.0761 1.2679 1.667 2.974
\end{Soutput}
\begin{Sinput}
R> namlab(dw)
\end{Sinput}
\begin{Soutput}
  Variable                  Label
1       id               Fruit Id
2      r_A  Fruit Radius - Apples
3      r_B Fruit Radius - Bananas
4      h_A  Fruit Height - Apples
5      h_B Fruit Height - Bananas
\end{Soutput}
\end{Schunk}
%
For the recast pivot (\code{how = "recast"}), unless a column named \code{variable} exists in the data, the source and (optionally) destination of variable names need to be specified using a list passed to \code{names}, and similarly for \code{labels}. Again, taking along labels is optional, and omitting either the list's \code{from} or \code{to} elements will omit the respective operations.
%
\begin{Schunk}
\begin{Sinput}
R> (dr <- pivot(data, ids = "id", names = list(from = "type"),
+               labels = list(from = "type_name", to = "label"), how = "r"))
\end{Sinput}
\begin{Soutput}
  id variable        label      A      B
1  1        r Fruit Radius 0.3916 0.4958
2  2        r Fruit Radius 1.0761 1.2679
3  1        h Fruit Height 1.8454 1.9762
4  2        h Fruit Height 1.6666 2.9737
\end{Soutput}
\begin{Sinput}
R> vlabels(dr)
\end{Sinput}
\begin{Soutput}
        id   variable      label          A          B 
"Fruit Id"         NA         NA   "Apples"  "Bananas" 
\end{Soutput}
\end{Schunk}
%
As with the other pivots, this operation does not incur any loss of information. A suitable reverse operation also exists: \code{pivot(dr, "id", names = list(to = "type"), labels = list(from = "label", to = "type_name"), how = "r")}. More features of \fct{pivot} are demonstrated in the \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}. Notably, it is also possible to perform longer and recast pivots without id variables. The recast pivot without ids resembles a generalization of \code{data.table::transpose()}, albeit slightly less efficient.
%
\section{List processing} \label{sec:list_proc}
%
Often in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects (such as a model object, performance statistics, and predictions) in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices, e.g., in response to nested user inputs in shiny apps. Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit. \newline

@@

        R 9.827 9.846 10.453  9.872 10.069 21.22   100
    Rfast 1.316 1.792  2.352  1.857  2.110 11.40   100
 collapse 1.324 1.449  1.965  1.520  1.772 11.31   100

@@

          R 10.921 11.244 11.810 11.315 11.673  16.293   100
 data.table  8.183 18.755 21.582 19.753 21.508 113.610   100
   collapse  1.913  2.819  3.393  3.051  3.445   7.233   100

@@

 data.table 135.72 137.46 139.66 138.75 140.71 158.89   100
   collapse  63.92  74.84  79.86  79.75  85.08  96.31   100

@@

 data.table 8.930 13.689 29.964 16.518 57.245 125.224   100
   collapse 1.281  1.419  1.553  1.485  1.585   2.295   100

@@

 data.table 11.65 15.19 22.12  16.39 18.31 62.51   100
   collapse 11.67 15.02 21.86  16.05 17.61 63.43   100

@@

 data.table 69.41 116.29 120.08 119.79 123.5 143.8   100
   collapse 57.53  88.93  92.62  91.98  96.6 125.0   100

@@

      base_int 59.641 62.592 64.06 63.488 64.871 111.96   100
  collapse_int  8.413  9.133 10.11  9.291  9.694  15.32   100
     base_char 91.712 93.905 96.68 95.830 97.707 141.16   100
 collapse_char 21.339 22.828 24.25 23.053 23.813  71.40   100

@@

        base_int 26.730 26.743 28.290  26.89 30.47 31.18    10
    collapse_int  8.584  8.609  9.103   8.66  8.69 13.19    10
       base_char 86.599 87.263 90.344  88.16 91.71 98.66    10
 data.table_char 41.636 41.751 42.335  41.83 41.88 47.11    10
   collapse_char 32.563 32.793 33.322  32.88 32.94 37.40    10

>>>>

\code{collapse::pivot} provides a modern class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings: it has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots, and supports complex labelled data without loss of information. Its syntax is:
%
\begin{Code}
pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
  how = "longer", na.rm = FALSE, factor = c("names", "labels"),
  check.dups = FALSE, nthreads = 1, fill = NULL, drop = TRUE,
  sort = FALSE, transpose = FALSE)
\end{Code}
%
The demonstration below employs a generated dataset about fruits. We could equivalently think about a survey with households and individuals, or sectors and firms. Variable labels are stored in \code{attr(column, "label")}. The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation} provides more elaborate examples.
%
\begin{Schunk}
\begin{Sinput}
R> data <- data.frame(type = rep(c("A", "B"), each = 2),
+    type_name = rep(c("Apples", "Bananas"), each = 2),
+    id = rep(1:2, 2), r = abs(rnorm(4)), h = abs(rnorm(4)*2))
R> setrelabel(data, id = "Fruit Id", r = "Fruit Radius", h = "Fruit Height")
R> print(data)
\end{Sinput}
\begin{Soutput}
  type type_name id      r      h
1    A    Apples  1 0.0356 2.4303
2    A    Apples  2 1.8575 0.5427
3    B   Bananas  1 0.5966 1.8641
4    B   Bananas  2 0.8115 0.7461
\end{Soutput}
\begin{Sinput}
R> vlabels(data)
\end{Sinput}
\begin{Soutput}
          type      type_name             id              r              h 
            NA             NA     "Fruit Id" "Fruit Radius" "Fruit Height" 
\end{Soutput}
\end{Schunk}
%
To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = c(...))}. If \code{labels = "lab_name"} is specified, variable labels are saved to an additional column named \code{lab\_name}. In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to assign alternative names to the \code{variable} and \code{value} columns, respectively.
%
\begin{Schunk}
\begin{Sinput}
R> (dl <- pivot(data, ids = c("type", "type_name", "id"), labels = "label"))
\end{Sinput}
\begin{Soutput}
  type type_name id variable        label  value
1    A    Apples  1        r Fruit Radius 0.0356
2    A    Apples  2        r Fruit Radius 1.8575
3    B   Bananas  1        r Fruit Radius 0.5966
4    B   Bananas  2        r Fruit Radius 0.8115
5    A    Apples  1        h Fruit Height 2.4303
6    A    Apples  2        h Fruit Height 0.5427
7    B   Bananas  1        h Fruit Height 1.8641
8    B   Bananas  2        h Fruit Height 0.7461
\end{Soutput}
\begin{Sinput}
R> vlabels(dl)
\end{Sinput}
\begin{Soutput}
      type  type_name         id   variable      label      value 
        NA         NA "Fruit Id"         NA         NA         NA 
\end{Soutput}
\end{Schunk}
%
\fct{pivot} only requires essential information and intelligently guesses the rest. For example, the same result could have been obtained by \code{pivot(data, values = c("r", "h"), labels = "label")}. An exact reverse operation can also be performed by specifying as little as \code{pivot(dl, labels = "label", how = "w")}. \newline

The second option is a wider pivot with \code{how = "wider"}. Here, \code{names} and \code{labels} can be used to select columns containing the names of new columns and their labels.\footnote{multiple columns with names and labels could be selected, which would be combined using \code{"\_"} for names and \code{" - "} for labels.} Note how the labels are combined with existing labels such that also this operation is without loss of information. It is, however, a destructive operation, i.e., with 2 or more columns selected through \code{values}, \fct{pivot} is not able to reverse it. Further arguments like \code{na.rm}, \code{fill}, \code{sort}, and \code{transpose} can be used to control the casting process.
%
\begin{Schunk}
\begin{Sinput}
R> (dw <- pivot(data, "id", names = "type", labels = "type_name", how = "w"))
\end{Sinput}
\begin{Soutput}
  id    r_A    r_B    h_A    h_B
1  1 0.0356 0.5966 2.4303 1.8641
2  2 1.8575 0.8115 0.5427 0.7461
\end{Soutput}
\begin{Sinput}
R> namlab(dw)
\end{Sinput}
\begin{Soutput}
  Variable                  Label
1       id               Fruit Id
2      r_A  Fruit Radius - Apples
3      r_B Fruit Radius - Bananas
4      h_A  Fruit Height - Apples
5      h_B Fruit Height - Bananas
\end{Soutput}
\end{Schunk}
%
For the recast pivot (\code{how = "recast"}), unless a column named \code{variable} exists in the data, the source and (optionally) destination of variable names need to be specified using a list passed to \code{names}, and similarly for \code{labels}. Again, taking along labels is optional, and omitting either the list's \code{from} or \code{to} elements will omit the respective operations.
%
\begin{Schunk}
\begin{Sinput}
R> (dr <- pivot(data, ids = "id", names = list(from = "type"),
+               labels = list(from = "type_name", to = "label"), how = "r"))
\end{Sinput}
\begin{Soutput}
  id variable        label      A      B
1  1        r Fruit Radius 0.0356 0.5966
2  2        r Fruit Radius 1.8575 0.8115
3  1        h Fruit Height 2.4303 1.8641
4  2        h Fruit Height 0.5427 0.7461
\end{Soutput}
\begin{Sinput}
R> vlabels(dr)
\end{Sinput}
\begin{Soutput}
        id   variable      label          A          B 
"Fruit Id"         NA         NA   "Apples"  "Bananas" 
\end{Soutput}
\end{Schunk}
%
As with the other pivots, this operation does not incur any loss of information. A suitable reverse operation also exists: \code{pivot(dr, "id", names = list(to = "type"), labels = list(from = "label", to = "type_name"), how = "r")}. More features of \fct{pivot} are demonstrated in the \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}. Notably, it is also possible to perform longer and recast pivots without id variables. The recast pivot without ids resembles a generalization of \code{data.table::transpose()}, albeit slightly less efficient.
%
\section{List processing} \label{sec:list_proc}
%
Often in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects (such as a model object, performance statistics, and predictions) in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices, e.g., in response to nested user inputs in shiny apps. Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit. \newline

@@

        R 9.755 9.871 10.279  9.978 10.422 14.36   100
    Rfast 1.374 1.823  2.460  1.974  2.376 15.22   100
 collapse 1.415 1.488  2.088  1.535  1.702 12.40   100

@@

          R 10.948 11.254 13.085 11.404 11.749 96.20   100
 data.table 13.031 19.024 22.010 19.743 23.276 62.05   100
   collapse  2.213  2.874  3.607  3.084  3.567 13.94   100

@@

 data.table 136.63 137.82 139.30 138.51 140.47 149.33   100
   collapse  73.28  79.03  82.12  81.74  84.15  99.39   100

@@

 data.table 8.525 12.288 24.414 14.395 20.293 73.406   100
   collapse 1.272  1.386  1.477  1.424  1.459  2.843   100

@@

 data.table 11.83 15.16 22.61  16.60 18.44 67.87   100
   collapse 11.75 14.87 22.46  16.33 18.52 71.10   100

@@

 data.table 73.64 113.04 117.94 115.83 121.94 181.0   100
   collapse 54.87  72.95  83.95  79.95  92.68 134.4   100

@@

      base_int 60.68 63.462 66.31 64.833  66.42 129.54   100
  collapse_int  8.50  9.193 10.48  9.374  10.23  19.99   100
     base_char 92.71 95.737 99.69 98.380 100.45 149.15   100
 collapse_char 22.25 23.781 25.86 24.143  25.14  79.24   100

@@

        base_int 27.156 27.313 29.131 27.956 31.399 32.45    10
    collapse_int  8.766  8.838  9.395  8.862  8.901 13.63    10
       base_char 84.300 84.414 88.440 85.985 89.799 98.14    10
 data.table_char 42.195 42.372 43.311 42.621 43.390 48.58    10
   collapse_char 29.627 29.786 30.585 30.038 30.367 34.70    10
